---
title: "Results"
author: "Marnin Wolfe"
date: "2020-Dec-03"
output: 
  workflowr::wflow_html:
    code_folding: hide
    toc: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = F, tidy = T)
```

# Cross-validation accuracy

Conducted 5-fold x 5-reps of cross-validation ([here](04-CrossValidation.html)). Three traits only, MCMDS, logFYLD, DM.

```{r}
library(tidyverse); library(magrittr);
cvresults<-readRDS(here::here("output","cvresults_ModelA_chunk1.rds")) %>% 
  bind_rows(readRDS(here::here("output","cvresults_ModelA_chunk2.rds"))) %>% 
  bind_rows(readRDS(here::here("output","cvresults_ModelA_chunk3.rds"))) %>% 
  mutate(Model="A") %>% 
  bind_rows(readRDS(here::here("output","cvresults_ModelADE_chunk1.rds")) %>% 
             bind_rows(readRDS(here::here("output","cvresults_ModelADE_chunk2.rds"))) %>% 
             bind_rows(readRDS(here::here("output","cvresults_ModelADE_chunk3.rds"))) %>% 
             mutate(Model="ADE"))
```

```{r}
cvresults %>% 
  select(Trait,repeats,id,VersionOfBLUPs,accGEBV,Model) %>% 
  ggplot(.,aes(x=Model,y=accGEBV,fill=VersionOfBLUPs)) + 
  geom_boxplot() +
  theme_bw() + facet_wrap(~Trait, scales = 'free') + 
  scale_fill_viridis_d() + labs(title="GEBV: Compare 3-stage and 2-stage prediction pipelines")
```

```{r}
cvresults %>% 
  select(Trait,repeats,id,VersionOfBLUPs,accGETGV,Model) %>% 
  ggplot(.,aes(x=Model,y=accGETGV,fill=VersionOfBLUPs)) + 
  geom_boxplot() +
  theme_bw() + facet_wrap(~Trait, scales = 'free') + 
  scale_fill_viridis_d() + labs(title="GETGV: Compare 3-stage and 2-stage prediction pipelines")
```

```{r}
cvresults %>% 
  select(Trait,Model,repeats,id,VersionOfBLUPs,accGEBV) %>% 
  spread(VersionOfBLUPs,accGEBV) %>% 
  mutate(diffAcc=blups2stage-blups3stage) %>% 
  ggplot(.,aes(x=Model,y=diffAcc,fill=Trait)) + 
  geom_hline(yintercept = 0, color='darkred') + 
  geom_boxplot() +
  theme_bw() + facet_wrap(~Trait,scales='free') + 
  scale_fill_viridis_d() + labs(y="Accuracy Difference (2-stage minus 3-stage)",title="GEBV")
```

```{r}
cvresults %>% 
  select(Trait,Model,repeats,id,VersionOfBLUPs,accGETGV) %>% 
  spread(VersionOfBLUPs,accGETGV) %>% 
  mutate(diffAcc=blups2stage-blups3stage) %>% 
  ggplot(.,aes(x=Model,y=diffAcc,fill=Trait)) + 
  geom_hline(yintercept = 0, color='darkred') + 
  geom_boxplot() +
  theme_bw() + facet_wrap(~Trait,scales='free') + 
  scale_fill_viridis_d() + labs(y="Accuracy Difference (2-stage minus 3-stage)",title="GETGV")
```

```{r}
cvresults %>% filter(VersionOfBLUPs=="blups2stage") %>% 
  select(Trait,repeats,id,VersionOfBLUPs,accGETGV,Model) %>% 
  ggplot(.,aes(x=Trait,y=accGETGV,fill=Model)) + 
  geom_boxplot(color='gray60', notch=T) +
  theme_bw() + facet_wrap(~Trait, scales = 'free') + 
  scale_fill_viridis_d() + labs(title="Compare accuracy: models A vs. ADE")
```

# Genetic Gain

## September 2020

```{r, fig.height=7, fig.width=5}
library(tidyverse); library(magrittr)
iita_gebvs<-read.csv(here::here("output","GEBV_IITA_ModelA_twostage_IITA_2020Sep21.csv"), stringsAsFactors = F)
traits<-c("DM","logFYLD","logTOPYLD","MCMDS")
iita_gebvs %>% 
  select(GID,GeneticGroup,any_of(traits)) %>% 
  pivot_longer(cols=any_of(traits),names_to = "Trait",values_to = "GEBV") %>% 
  group_by(Trait,GeneticGroup) %>% 
  summarize(meanGEBV=mean(GEBV),
            stdErr=sd(GEBV)/sqrt(n()),
            upperSE=meanGEBV+stdErr,
            lowerSE=meanGEBV-stdErr) %>% 
  ggplot(.,aes(x=GeneticGroup,y=meanGEBV,fill=Trait)) + 
  geom_bar(stat = 'identity', color='gray60', size=1.25) + 
  geom_linerange(aes(ymax=upperSE,
                     ymin=lowerSE), color='gray60', size=1.25) + 
  facet_wrap(~Trait,scales='free_y', ncol=1) + 
  theme_bw() +
  geom_hline(yintercept = 0, size=1.15, color='black') + 
  theme(axis.text.x = element_text(face = 'bold',angle = 0, size=12),
        axis.title.y = element_text(face = 'bold',size=14),
        legend.position = 'none',
        strip.background.x = element_blank(),
        strip.text = element_text(face='bold',size=14)) + 
  scale_fill_viridis_d() + 
  labs(x=NULL,y="Mean GEBVs")
```

# Rate of gain {#rate_of_gain}

```{r, eval = F}
# List of trials from 2020 to Prasad and Ismail... should I download fresh data?
# dbdata<-readRDS(here::here("output","IITA_CleanedTrialData.rds"))
# trialsHarvested2019to2020<-dbdata %>% 
#   filter(studyYear>=2019) %>% 
#   group_by(studyYear,locationName,studyName,plantingDate,harvestDate) %>% 
#   summarize(Nhav=sum(!is.na(NOHAV)))
# trialsHarvested2019to2020 %>% write.csv(.,file=here::here("output","trials_uploaded_by_Nharvested_15Sep2020.csv"), row.names=F)
```

## GETGV vs. "Accession Year"

Start by merging the "accession year" variable with the GETGVs.

```{r}
library(tidyverse); library(magrittr);
iita_getgvs<-read.csv(here::here("output","GETGV_IITA_ModelADE_twostage_IITA_2020Dec03.csv"), stringsAsFactors = F)
traits<-c("logDYLD","logFYLD","MCMDS","DM","BCHROMO",
          "BRLVLS","HI","logTOPYLD")
# traits<-c("MCMDS","DM","PLTHT","BRNHT1","BRLVLS","HI",
#           "logDYLD", "logFYLD","logTOPYLD","logRTNO","TCHART","LCHROMO","ACHROMO","BCHROMO")
ggcycletime<-readxl::read_xls(here::here("data","PedigreeGeneticGainCycleTime_aafolabi_01122020.xls"))
# table(ggcycletime$Accession %in% iita_getgvs$GID)
# FALSE 
#   807 
# Need germplasmName field from raw trial data to match GEBV and cycle time
dbdata<-readRDS(here::here("output","IITA_ExptDesignsDetected_2020Dec03.rds"))
iita_getgvs %<>% 
  left_join(dbdata %>% 
  select(-MaxNOHAV) %>% unnest(TrialData) %>% 
  distinct(germplasmName,GID)) %>% 
  group_by(GID) %>% 
  slice(1) %>% 
  ungroup()
rm(dbdata)
# table(ggcycletime$Accession %in% iita_getgvs$germplasmName)
# FALSE  TRUE 
#   193   614 

# table(ggcycletime$Year_Accession)
iita_getgvs %<>% 
  left_join(.,ggcycletime %>% 
              rename(germplasmName=Accession) %>% 
              mutate(Year_Accession=as.numeric(Year_Accession)))
iita_getgvs %<>% 
  mutate(Year_Accession=case_when(grepl("2013_|TMS13",germplasmName)~2013,
                                  grepl("TMS14",germplasmName)~2014,
                                  grepl("TMS15",germplasmName)~2015,
                                  grepl("TMS18",germplasmName)~2018,
                                  !grepl("2013_|TMS13|TMS14|TMS15|TMS18",germplasmName)~Year_Accession)) 

write.csv(iita_getgvs, file = here::here("output","GETGV_IITA_ModelADE_twostage_IITA_2020Dec03_withAccessionYear.csv"), row.names = F)

```

Key output is a file **`output/GETGV_IITA_ModelADE_twostage_IITA_2020Dec03_withAccessionYear.csv`** for use in downstream analyses.

## What is yellow?

```{r}
rm(list=ls())
library(tidyverse); library(magrittr);
iita_getgvs<-read.csv(here::here("output",
                                 "GETGV_IITA_ModelADE_twostage_IITA_2020Dec03_withAccessionYear.csv"),
                      stringsAsFactors = F)

traits<-c("logDYLD","logFYLD","MCMDS","DM","BCHROMO",
          "BRLVLS","HI","logTOPYLD")
```

Plot B-value and decide on a threshold for removing "yellow" clones from the analysis.

```{r, fig.width=10, fig.height=5}
# iita_getgvs %>% 
#   ggplot(.,aes(x=TCHART,y=BCHROMO)) + geom_hex() +
#   theme_bw() + facet_wrap(~GeneticGroup, nrow=1) + theme(legend.position = 'none') + 
#   geom_vline(xintercept = 0.5) + geom_hline(yintercept = 5) + 
#   labs(title="Arbitrary suggested cut-offs for `white` rooted GETGVs", subtitle = "horiz. and vert. lines")
```

```{r, fig.width=10, fig.height=5}
iita_getgvs %>% 
  ggplot(.,aes(x=BCHROMO)) + geom_histogram() +
  theme_bw() + #facet_wrap(~GeneticGroup, nrow=1) + 
  theme(legend.position = 'none') + 
  geom_vline(xintercept = 1, color='darkred') +  # geom_hline(yintercept = 5) + 
  labs(title="Histogram of GETGV for chromometer B-value",
       subtitle = "Cut-offs for `white` roots == 1")
```

## Subset years

Remove clones between 2005 and 2012.

Declare the "eras" as PreGS\<2012 and GS\>=2013.

```{r}
iita_getgvs %<>% 
  filter(Year_Accession>2012 | Year_Accession<2005)

iita_getgvs %<>% 
  mutate(GeneticGroup=ifelse(Year_Accession>=2013,"GS","PreGS"))
```

## Analysis by raw GETGVs

Number of clones for each "era" that

```{r}
iita_getgvs %>% 
  select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
  count(Nclone=GeneticGroup)
```

Number of white root clones (`BCHROMO<=1`).

```{r}
iita_getgvs %>% 
  select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
  filter(BCHROMO<=1) %>% 
  count(NwhiteRoot=GeneticGroup)
```

Group by Era (Genetic Group) and fit a simple linear regression for each trait, i.e. `lm(GETGV ~ Year_Accession)`.

Fit model to "all clones" and then to "white root clones only".

```{r}
model_rawgetgvs<-iita_getgvs %>% 
  select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
  mutate(Dataset="AllGermplasm") %>% 
  bind_rows(iita_getgvs %>% 
              select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
              filter(BCHROMO<=1) %>% 
              mutate(Dataset="WhiteRootClones")) %>% 
  pivot_longer(cols = all_of(traits), names_to = "Trait", values_to = "GETGV") %>% 
  nest(data=c(GID,Year_Accession,GETGV)) %>% 
  mutate(model=map(data,~lm(formula = "GETGV~Year_Accession", data = .)))
```

Extract the model effects, etc.

```{r}
model_rawgetgvs %<>% 
  mutate(out=map(model,~broom::glance(.))) %>% 
  unnest(out)
model_rawgetgvs %<>% 
  mutate(out=map(model,~broom::tidy(.)))
model_rawgetgvs %<>% 
  mutate(out=map(out,~select(.,term,estimate) %>% spread(term,estimate))) %>% 
  unnest(out) %>% 
  rename(InterceptEst=`(Intercept)`,
         YearAccessionEst=Year_Accession) %>% 
  select(Dataset,GeneticGroup,Trait,r.squared,nobs,InterceptEst,YearAccessionEst)
```

Basic summary of linear models

```{r, rows.print=32}
model_rawgetgvs %>% rmarkdown::paged_table()
```

Compare slope estimates between "eras"

```{r, rows.print=16}
model_rawgetgvs %>% 
  select(Dataset,GeneticGroup,Trait,YearAccessionEst) %>% 
  spread(GeneticGroup,YearAccessionEst) %>% 
  rmarkdown::paged_table()
```

Add some summary of the raw data that went into the GETGV analyzed above.

```{r}
# summarize the raw plot data
dbdata<-readRDS(here::here("output","IITA_ExptDesignsDetected_2020Dec03.rds")) %>% 
  dplyr::select(-MaxNOHAV) %>% 
  unnest(TrialData) %>% 
  dplyr::select(programName,locationName,studyYear,TrialType,studyName,
         CompleteBlocks,IncompleteBlocks,
         yearInLoc,trialInLocYr,repInTrial,blockInRep,observationUnitDbId,
         germplasmName,FullSampleName,GID,all_of(traits),PropNOHAV) %>% 
  mutate(IncompleteBlocks=ifelse(IncompleteBlocks==TRUE,"Yes","No"),
         CompleteBlocks=ifelse(CompleteBlocks==TRUE,"Yes","No")) %>% 
  pivot_longer(cols = all_of(traits), names_to = "Trait", values_to = "Value") %>% 
  filter(!is.na(Value),
         !is.na(GID)) %>% 
  nest(MultiTrialTraitData=c(-Trait))

trainingdata_summary<-dbdata %>% 
  mutate(NplotsTotal=map_dbl(MultiTrialTraitData,nrow),
         nplot=map(MultiTrialTraitData,~count(.,TrialType) %>% mutate(TrialType=paste0("Nplots_",TrialType)) %>% 
                     spread(TrialType,n) %>% 
                     select(any_of(paste0("Nplots_",c("CrossingBlock","GeneticGain","CET","ExpCET","PYT","AYT","UYT","NCRP")))))) %>% 
  unnest(nplot) %>% 
  select(-MultiTrialTraitData) %>% 
  # and add a summary of the BLUPs that result
  # which were then later used for prediction
  left_join(readRDS(file=here::here("output","iita_blupsForModelTraining_twostage_asreml_2020Dec03.rds")) %>% 
              filter(Trait %in% traits) %>% 
              mutate(NclonesWithBLUPs=map_dbl(blups,nrow)) %>% 
              select(Trait,NclonesWithBLUPs,Vg,Ve,H2))
```

Print a summary of the raw plots and resulting BLUPs that went into the GETGV .

```{r, rows.print=8, cols.print=14}
trainingdata_summary %>% rmarkdown::paged_table()
```

Write model summaries to disk: **`output/model_rawgetgvs_vs_year.csv`**.

Write training data summary to disk: **`output/training_data_summary.csv`**

```{r}
write.csv(trainingdata_summary,file = here::here("output","training_data_summary.csv"), row.names = F)
write.csv(model_rawgetgvs,file = here::here("output","model_rawgetgvs_vs_year.csv"), row.names = F)
```

### Plot all germplasm vs. year

```{r, fig.height=10, fig.width=12}
iita_getgvs %>% 
  select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
  pivot_longer(cols = all_of(traits), names_to = "Trait", values_to = "GETGV") %>% 
  mutate(Trait=factor(Trait,traits)) %>% 
  ggplot(.,aes(x=Year_Accession,y=GETGV,color=GeneticGroup)) + 
  geom_point(size=1.25) + geom_smooth(method=lm, se=TRUE, size=1.5) + 
  facet_wrap(~Trait,scales='free_y', ncol=2) + 
  theme_bw() +
  theme(axis.text = element_text(face = 'bold',angle = 0, size=14),
        axis.title = element_text(face = 'bold',size=16),
        strip.background.x = element_blank(),
        strip.text = element_text(face='bold',size=18)) + 
  scale_color_viridis_d() + 
  labs(title = "Regression of raw GETGV vs. Year_Accession by 'era' [GS vs. PreGS]",
       subtitle = "All Germplasm")
```

```{r, fig.height=10, fig.width=12}
iita_getgvs %>% 
  select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
  filter(BCHROMO<=1) %>% 
  pivot_longer(cols = all_of(traits), names_to = "Trait", values_to = "GETGV") %>% 
  mutate(Trait=factor(Trait,traits)) %>% 
  ggplot(.,aes(x=Year_Accession,y=GETGV,color=GeneticGroup)) + 
  geom_point(size=1.25) + geom_smooth(method=lm, se=TRUE, size=1.5) + 
  facet_wrap(~Trait,scales='free_y', ncol=2) + 
  theme_bw() +
  theme(axis.text = element_text(face = 'bold',angle = 0, size=14),
        axis.title = element_text(face = 'bold',size=16),
        strip.background.x = element_blank(),
        strip.text = element_text(face='bold',size=18)) + 
  scale_color_viridis_d() + 
  labs(title = "Regression of raw GETGV vs. Year_Accession by 'era' [GS vs. PreGS]",
       subtitle = "White Root Germplasm (BCHROMO<=1)")
```

## Analysis by mean GETGV-by-Year

**I recommend using the analysis and maybe also the plots above.**

For completeness, below is an analysis and plots using the meanGETGV-by-year.

Compute mean and std. error by Dataset ("all germplasm" vs. "white root clones only") and GeneticGroup ("GS" vs. "PreGS").

```{r}
mean_getgvs<-iita_getgvs %>% 
  select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
  mutate(Dataset="AllGermplasm") %>% 
  bind_rows(iita_getgvs %>% 
              select(GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
              filter(BCHROMO<=1) %>% 
              mutate(Dataset="WhiteRootClones")) %>%  
  select(Dataset,GeneticGroup,GID,Year_Accession,all_of(traits)) %>% 
  pivot_longer(cols = all_of(traits), names_to = "Trait", values_to = "GETGV") %>% 
  group_by(Dataset,Trait,GeneticGroup,Year_Accession) %>% 
  summarize(meanGETGV=mean(GETGV),
            Nclones=n(),
            stdErr=sd(GETGV)/sqrt(n()),
            upperSE=meanGETGV+stdErr,
            lowerSE=meanGETGV-stdErr) %>% 
  ungroup()

write.csv(mean_getgvs, file = here::here("output","meanGETGVbyYear_IITA_2020Dec03.csv"), row.names = F)
```

Group by Era (Genetic Group) and fit a simple linear regression for each trait, i.e. `lm(GETGV ~ Year_Accession)`.

```{r}
model_meangetgvs<-mean_getgvs %>% 
  nest(data=c(-Dataset,-Trait,-GeneticGroup)) %>% 
  mutate(model=map(data,~lm(formula = "meanGETGV~Year_Accession", data = .)))
```

Extract the model effects, etc.

```{r}
model_meangetgvs %<>% 
  mutate(out=map(model,~broom::glance(.))) %>% 
  unnest(out) %>% 
  mutate(out=map(model,~broom::tidy(.))) %>% 
  mutate(out=map(out,~select(.,term,estimate) %>% spread(term,estimate))) %>% 
  unnest(out) %>% 
  rename(InterceptEst=`(Intercept)`,
         YearAccessionEst=Year_Accession) %>% 
  select(Dataset,GeneticGroup,Trait,r.squared,nobs,InterceptEst,YearAccessionEst)
```

Basic summary of linear models

```{r, rows.print=32}
model_meangetgvs %>% rmarkdown::paged_table()
```

Compare slope estimates between "eras"

```{r, rows.print=16}
model_meangetgvs %>% 
  select(Dataset,GeneticGroup,Trait,YearAccessionEst) %>% 
  spread(GeneticGroup,YearAccessionEst) %>% 
  rmarkdown::paged_table()
```

Save these estimates also to disk at: **`output/model_meangetgvs_vs_year.csv`**

```{r}
write.csv(model_meangetgvs,file = here::here("output","model_meangetgvs_vs_year.csv"), row.names = F)
```

### Plot all germplasm vs. year

```{r, fig.height=10, fig.width=12}
mean_getgvs %>% 
  filter(Dataset=="AllGermplasm") %>% 
  mutate(Trait=factor(Trait,traits)) %>% 
  ggplot(.,aes(x=Year_Accession,y=meanGETGV,color=GeneticGroup,size=Nclones)) + 
  geom_point(size=4) + geom_smooth(method=lm, se=TRUE) + 
  geom_linerange(aes(ymax=upperSE,
                     ymin=lowerSE), color='gray40', size=1) + 
  facet_wrap(~Trait,scales='free_y', ncol=2) + 
  theme_bw() +
  theme(axis.text = element_text(face = 'bold',angle = 0, size=14),
        axis.title = element_text(face = 'bold',size=16),
        strip.background.x = element_blank(),
        strip.text = element_text(face='bold',size=18)) + 
  scale_color_viridis_d() + 
  labs(title = "meanGETGV vs. Year_Accession - All Germplasm",
       subtitle = "Mean across all clones within Year_Accession")
```

### Plot "white" germplasm vs. year

```{r, fig.height=10, fig.width=12}
mean_getgvs %>% 
  filter(Dataset=="WhiteRootClones") %>% 
  mutate(Trait=factor(Trait,traits)) %>% 
  ggplot(.,aes(x=Year_Accession,y=meanGETGV,color=GeneticGroup,size=Nclones)) + 
  geom_point(size=4) + geom_smooth(method=lm, se=TRUE) + 
  geom_linerange(aes(ymax=upperSE,
                     ymin=lowerSE), color='gray40', size=1) + 
  facet_wrap(~Trait,scales='free_y', ncol=2) + 
  theme_bw() +
  theme(axis.text = element_text(face = 'bold',angle = 0, size=14),
        axis.title = element_text(face = 'bold',size=16),
        strip.background.x = element_blank(),
        strip.text = element_text(face='bold',size=18)) + 
  scale_color_viridis_d() + 
  labs(title = "meanGETGV vs. Year_Accession - White rooted germplasm",
       subtitle = "Mean across all clones within Year_Accession")
```
