---
title: "Review and QC of IITA training data"
author: "Marnin Wolfe"
date: "2020-Dec-03"
output: 
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = F, tidy = T)
```

Follow outlined GenomicPredictionChecklist and previous pipeline to process cassavabase data for ultimate genomic prediction.

Below we will clean and format training data.

* Inputs: "Raw" field trial data
* Expected outputs: "Cleaned" field trial data

# Re-fresh data from Sept. '20

This will update genomic predictions relative to the [earlier analysis done in Sept. 2020](01-cleanTPdata.html). 

# Initial cassavabase download

## [User input] Cassavabase download

Downloaded all IITA field trials with **studyYear** 2018, 2019, 2020.

1. [Cassavabase search wizard](https://www.cassavabase.org/breeders/search):  
2. Selected all IITA trials with **studyYear** 2018, 2019, 2020. Make a list. Named it **IITA_Trials_2018to2020_2020Dec03**.
3. Go to **Manage** --> **Download** [here](https://www.cassavabase.org/breeders/download). Download phenotypes (plot-basis only) and meta-data as CSV using the corresponding boxes / drop-downs. 
4. Store flatfiles, unaltered in directory [`DatabaseDownload_2020Dec03/` uploaded to Cassavabase FTP server](ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/DatabaseDownload_2020Sep15/). 

**2018 trials:** probably redundant to those previously [downloaded in July 2019](ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/DatabaseDownload_72419/) for the [genomic prediction of GS C4](https://wolfemd.github.io/IITA_2019GS/). In case some trials weren't harvested as of July 2019, use the 2018 trials downloaded here instead of the ones from 2019.

**2019 trials:** All trials harvested and uploaded as of now (Dec. 3rd, 2020) are to be added to refresh the genomic predictions. 

**2020 trials:** If any current trials already have e.g. disease data, will use it.

# [User input] Cassavabase download

Downloaded **all** IITA field trials.

1. [Cassavabase search wizard](https://www.cassavabase.org/breeders/search):  
2. Selected *all* IITA trials currently available. Make a list. Named it **ALL_IITA_TRIALS_2020Sep15**.
3. Go to **Manage** --> **Download** [here](https://www.cassavabase.org/breeders/download). Download phenotypes (plot-basis only) and meta-data as CSV using the corresponding boxes / drop-downs. 
4. Store flatfiles, unaltered in directory [`DatabaseDownload_2020Sep15/` uploaded to Cassavabase FTP server](ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/DatabaseDownload_2020Sep15/). 

## Join metadata files

**Possible database bug?** The entire >500Mb phenotype dataset for IITA downloaded without a problem. However, I'm getting an "server error" message trying to download the corresponding meta-data in one chunk. 

**Solution:** combine meta-data downloaded for "all" trials in July 2019, with meta-data download for the 2018-2020 period done Dec 3rd, 2020. Feed joined file to `readDBdata()`.

```{r, eval=F}
metadata19<-read.csv("ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/DatabaseDownload_72419/2019-07-24T144144metadata_download.csv", 
                     na.strings = c("#VALUE!",NA,".",""," ","-","\""),
                     stringsAsFactors = F)
metadata20<-read.csv("ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/DatabaseDownload_2020Dec03/2020-12-03T094057metadata_download.csv",
                     na.strings = c("#VALUE!",NA,".",""," ","-","\""),
                     stringsAsFactors = F)
metadata19 %>% 
  # remove lines for trials in the 2020 download
  filter(!studyName %in% metadata20$studyName) %>% 
  bind_rows(metadata20) %>% 
  # ensure no duplicate lines
  distinct %>% 
  # write to disk
  write.csv(.,here::here("output","all_iita_metadata_Dec2020.csv"),row.names = F)
```

# Read-in trial data

```{r, message=F}
rm(list=ls())
library(tidyverse); library(magrittr);
source(here::here("code","gsFunctions.R"))
```

Read DB data directly from the Cassavabase FTP server.
```{r}
## From September
dbdata_sep2020<-readDBdata(phenotypeFile = paste0("ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/",
                                                  "DatabaseDownload_2020Sep15/2020-09-15T185453phenotype_download.csv"),
                           metadataFile = here::here("output","all_iita_metadata.csv"))
## New, December download
dbdata_dec2020<-readDBdata(phenotypeFile = paste0("ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/",
                                                  "DatabaseDownload_2020Dec03/2020-12-03T102130phenotype_download.csv"),
                           metadataFile = here::here("output","all_iita_metadata_Dec2020.csv"))
# dbdata_dec2020 %>% 
#   filter(!is.na(plantingDate)) %>% count(studyName,studyYear,locationName)
#dbdata_dec2020 %>% select(studyYear,studyDbId)
# table(unique(dbdata_dec2020$studyName) %in% unique(dbdata_sep2020$studyName))
# FALSE  TRUE 
#    42   371 

# 42 "new trials"
#table(dbdata_dec2020$observationUnitDbId %in% dbdata_sep2020$observationUnitDbId)
 # FALSE   TRUE 
 #  5823 175964 

# length(dbdata_dec2020$observationUnitDbId)==length(unique(dbdata_dec2020$observationUnitDbId))

# any of the "new" plots register as being for trials already in the sep data?
### yes, 1 trial (see below)
# table(unique(dbdata_dec2020$studyName[!dbdata_dec2020$observationUnitDbId %in% dbdata_sep2020$observationUnitDbId]) %in% unique(dbdata_sep2020$studyName))
# FALSE  TRUE 
#    42     1 

### probably doesn't matter, just replace the old with the new

dbdata<-bind_rows(dbdata_sep2020 %>% 
                    mutate(replicate=as.integer(replicate),
                           rowNumber=as.integer(rowNumber),
                           colNumber=as.integer(colNumber)) %>% 
                    mutate(across(contains("CO_334"),as.numeric)) %>% 
                    anti_join(dbdata_dec2020 %>% 
                                distinct(observationUnitDbId,studyName,studyYear,studyDbId)),
                  dbdata_dec2020)
```
# Group and select trials to analyze

Make TrialType Variable

```{r}
dbdata<-makeTrialTypeVar(dbdata) 
dbdata %>% 
  count(TrialType)
```

## Trials NOT included

Looking at the **studyName**'s of trials getting NA for TrialType, which can't be classified at present.

Here is the list of trials I am _not_ including.
```{r}
dbdata %>% filter(is.na(TrialType)) %$% unique(studyName) %>% 
  write.csv(.,file = here::here("output","iita_trials_NOT_identifiable_Dec2020.csv"), row.names = F)
```
Wrote to disk a CSV in the `output/` sub-directory.

Should any of these trials have been included?

Especially the following new trials (post 2018)?
```{r}
dbdata %>% 
  filter(is.na(TrialType),
         as.numeric(studyYear)>2018) %$% unique(studyName)
```

## Remove unclassified trials

```{r}
dbdata %<>% 
    filter(!is.na(TrialType)) 
dbdata %>% 
    group_by(programName) %>% 
    summarize(N=n())
#	475097 plots (~155K are seedling nurseries which will be excluded from most analyses)
```

# Traits and TraitAbbreviations

Making a table of abbreviations for renaming. Since July 2019 version: added chromometer traits (L, a, b) and added branching levels count (BRLVLS) at IYR's request. 
```{r}
traitabbrevs<-tribble(~TraitAbbrev,~TraitName,
        "CMD1S","cassava.mosaic.disease.severity.1.month.evaluation.CO_334.0000191",
        "CMD3S","cassava.mosaic.disease.severity.3.month.evaluation.CO_334.0000192",
        "CMD6S","cassava.mosaic.disease.severity.6.month.evaluation.CO_334.0000194",
        "CMD9S","cassava.mosaic.disease.severity.9.month.evaluation.CO_334.0000193",
        "CGM","Cassava.green.mite.severity.CO_334.0000033",
        "CGMS1","cassava.green.mite.severity.first.evaluation.CO_334.0000189",
        "CGMS2","cassava.green.mite.severity.second.evaluation.CO_334.0000190",
        "DM","dry.matter.content.percentage.CO_334.0000092",
        "PLTHT","plant.height.measurement.in.cm.CO_334.0000018",
        "BRNHT1","first.apical.branch.height.measurement.in.cm.CO_334.0000106",
        "BRLVLS","branching.level.counting.CO_334.0000079",
        "SHTWT","fresh.shoot.weight.measurement.in.kg.per.plot.CO_334.0000016",
        "RTWT","fresh.storage.root.weight.per.plot.CO_334.0000012",
        "RTNO","root.number.counting.CO_334.0000011",
        "TCHART","total.carotenoid.by.chart.1.8.CO_334.0000161",
        "LCHROMO","L.chromometer.value.CO_334.0002065",
        "ACHROMO","a.chromometer.value.CO_334.0002066",
        "BCHROMO","b.chromometer.value.CO_334.0002064",
        "NOHAV","plant.stands.harvested.counting.CO_334.0000010")
traitabbrevs %>% rmarkdown::paged_table()
```

Run function `renameAndSelectCols()` to rename columns and remove everything unecessary

```{r}
dbdata<-renameAndSelectCols(traitabbrevs,indata=dbdata,customColsToKeep = "TrialType")
```

# QC Trait values

Standard code, recycled... should be a function?
```{r}
dbdata<-dbdata %>% 
  mutate(CMD1S=ifelse(CMD1S<1 | CMD1S>5,NA,CMD1S),
         CMD3S=ifelse(CMD3S<1 | CMD3S>5,NA,CMD3S),
         CMD6S=ifelse(CMD6S<1 | CMD1S>5,NA,CMD6S),
         CMD9S=ifelse(CMD9S<1 | CMD1S>5,NA,CMD9S),
         CGM=ifelse(CGM<1 | CGM>5,NA,CGM),
         CGMS1=ifelse(CGMS1<1 | CGMS1>5,NA,CGMS1),
         CGMS2=ifelse(CGMS2<1 | CGMS2>5,NA,CGMS2),
         DM=ifelse(DM>100 | DM<=0,NA,DM),
         RTWT=ifelse(RTWT==0 | NOHAV==0 | is.na(NOHAV),NA,RTWT),
         SHTWT=ifelse(SHTWT==0 | NOHAV==0 | is.na(NOHAV),NA,SHTWT),
         RTNO=ifelse(RTNO==0 | NOHAV==0 | is.na(NOHAV),NA,RTNO),
         NOHAV=ifelse(NOHAV==0,NA,NOHAV),
         NOHAV=ifelse(NOHAV>42,NA,NOHAV),
         RTNO=ifelse(!RTNO %in% 1:10000,NA,RTNO))
```

# Post-QC traits

## Harvest index
```{r}
dbdata<-dbdata %>% 
    mutate(HI=RTWT/(RTWT+SHTWT))
```

## Unit area traits

I anticipate this will not be necessary as it will be computed before or during data upload.

For calculating fresh root yield: 

1. **PlotSpacing:** Area in $m^2$ per plant. plotWidth and plotLength metadata would hypothetically provide this info, but is missing for vast majority of trials. Therefore, use info from Fola.
2. **maxNOHAV:** Instead of ExpectedNOHAV. Need to know the max number of plants in the area harvested. For some trials, only the inner (or "net") plot is harvested, therefore the PlantsPerPlot meta-variable will not suffice. Besides, the PlantsPerPlot information is missing for the vast majority of trials. Instead, use observed max(NOHAV) for each trial. We use this plus the PlotSpacing to calc. the area over which the RTWT was measured. During analysis, variation in the actual number of plants harvested will be accounted for.

```{r, message=F, warning=F}
dbdata<-dbdata %>% 
    mutate(PlotSpacing=ifelse(programName!="IITA",1,
                              ifelse(studyYear<2013,1,
                              ifelse(TrialType %in% c("CET","GeneticGain","ExpCET"),1,0.8))))
maxNOHAV_byStudy<-dbdata %>% 
  group_by(programName,locationName,studyYear,studyName,studyDesign) %>% 
  summarize(MaxNOHAV=max(NOHAV, na.rm=T)) %>% 
  ungroup() %>% 
  mutate(MaxNOHAV=ifelse(MaxNOHAV=="-Inf",NA,MaxNOHAV))

write.csv(maxNOHAV_byStudy %>% arrange(studyYear),file=here::here("output","maxNOHAV_byStudy.csv"), row.names = F)
```

```{r}
# I log transform yield traits 
# to satisfy homoskedastic residuals assumption 
# of linear mixed models
dbdata<-left_join(dbdata,maxNOHAV_byStudy) %>% 
  mutate(RTWT=ifelse(NOHAV>MaxNOHAV,NA,RTWT),
         SHTWT=ifelse(NOHAV>MaxNOHAV,NA,SHTWT),
         RTNO=ifelse(NOHAV>MaxNOHAV,NA,RTNO),
         HI=ifelse(NOHAV>MaxNOHAV,NA,HI),
         FYLD=RTWT/(MaxNOHAV*PlotSpacing)*10,
         DYLD=FYLD*(DM/100),
         logFYLD=log(FYLD),
         logDYLD=log(DYLD),
         logTOPYLD=log(SHTWT/(MaxNOHAV*PlotSpacing)*10),
         logRTNO=log(RTNO),
         PropNOHAV=NOHAV/MaxNOHAV) 
# remove non transformed / per-plot (instead of per area) traits
dbdata %<>% select(-RTWT,-SHTWT,-RTNO,-FYLD,-DYLD)
```

## Season-wide mean CMDS
```{r}
dbdata<-dbdata %>% 
  mutate(MCMDS=rowMeans(.[,c("CMD1S","CMD3S","CMD6S","CMD9S")], na.rm = T)) %>% 
  select(-CMD1S,-CMD3S,-CMD6S,-CMD9S)
```

# [User input] Assign genos to phenos

This step is mostly copy-pasted from previous processing of IITA-specific data. 

Uses 3 flat files, which are available e.g. [here](ftp://ftp.cassavabase.org/marnin_datasets/NRCRI_2020GS/data/). Specifically, `IITA_GBStoPhenoMaster_33018.csv`, `GBSdataMasterList_31818.csv` and `NRCRI_GBStoPhenoMaster_40318.csv`. I copy them to the `data/` sub-directory for the current analysis.

In addition, DArT-only samples are now expected to also have phenotypes. Therefore, checking for matches in new flatfiles, deposited in the `data/` (see code below).

```{r}
library(tidyverse); library(magrittr)
gbs2phenoMaster<-dbdata %>% 
  select(germplasmName) %>% 
  distinct %>% 
  left_join(read.csv(here::here("data","NRCRI_GBStoPhenoMaster_40318.csv"), 
                     stringsAsFactors = F)) %>% 
  mutate(FullSampleName=ifelse(grepl("C2a",germplasmName,ignore.case = T) & 
                                 is.na(FullSampleName),germplasmName,FullSampleName)) %>% 
  filter(!is.na(FullSampleName)) %>% 
  select(germplasmName,FullSampleName) %>% 
  bind_rows(dbdata %>% 
              select(germplasmName) %>% 
              distinct %>% 
              left_join(read.csv(here::here("data","IITA_GBStoPhenoMaster_33018.csv"), 
                                 stringsAsFactors = F)) %>% 
              filter(!is.na(FullSampleName)) %>% 
              select(germplasmName,FullSampleName)) %>% 
  bind_rows(dbdata %>% 
              select(germplasmName) %>% 
              distinct %>% 
              left_join(read.csv(here::here("data","GBSdataMasterList_31818.csv"), 
                                 stringsAsFactors = F) %>% 
                          select(DNASample,FullSampleName) %>% 
                          rename(germplasmName=DNASample)) %>% 
              filter(!is.na(FullSampleName)) %>% 
              select(germplasmName,FullSampleName)) %>% 
  bind_rows(dbdata %>% 
              select(germplasmName) %>% 
              distinct %>% 
              mutate(germplasmSynonyms=ifelse(grepl("^UG",germplasmName,ignore.case = T),
                                              gsub("UG","Ug",germplasmName),germplasmName)) %>% 
              left_join(read.csv(here::here("data","GBSdataMasterList_31818.csv"), 
                                 stringsAsFactors = F) %>% 
                          select(DNASample,FullSampleName) %>% 
                          rename(germplasmSynonyms=DNASample)) %>% 
              filter(!is.na(FullSampleName)) %>% 
              select(germplasmName,FullSampleName)) %>%  
  bind_rows(dbdata %>% 
              select(germplasmName) %>% 
              distinct %>% 
              mutate(germplasmSynonyms=ifelse(grepl("^TZ",germplasmName,
                                                    ignore.case = T),
                                              gsub("TZ","",germplasmName),germplasmName)) %>% 
              left_join(read.csv(here::here("data","GBSdataMasterList_31818.csv"), 
                                 stringsAsFactors = F) %>% 
                          select(DNASample,FullSampleName) %>% 
                          rename(germplasmSynonyms=DNASample)) %>% 
              filter(!is.na(FullSampleName)) %>%
              select(germplasmName,FullSampleName)) %>% 
  distinct %>% 
  left_join(read.csv(here::here("data","GBSdataMasterList_31818.csv"), 
                     stringsAsFactors = F) %>% 
              select(FullSampleName,OrigKeyFile,Institute) %>% 
              rename(OriginOfSample=Institute)) %>% 
  mutate(OrigKeyFile=ifelse(grepl("C2a",germplasmName,ignore.case = T),
                            ifelse(is.na(OrigKeyFile),"LavalGBS",OrigKeyFile),
                            OrigKeyFile),
         OriginOfSample=ifelse(grepl("C2a",germplasmName,ignore.case = T),
                               ifelse(is.na(OriginOfSample),"NRCRI",OriginOfSample),
                               OriginOfSample))
## NEW: check for germName-DArT name matches
germNamesWithoutGBSgenos<-dbdata %>% 
  select(programName,germplasmName) %>% 
  distinct %>% 
  left_join(gbs2phenoMaster) %>% 
  filter(is.na(FullSampleName)) %>% 
  select(-FullSampleName)

germNamesWithDArT<-germNamesWithoutGBSgenos %>% 
  inner_join(read.table(here::here("data","chr1_RefPanelAndGSprogeny_ReadyForGP_72719.fam"), header = F, stringsAsFactors = F)$V2 %>% 
               grep("TMS16|TMS17|TMS18|TMS19|TMS20",.,value = T, ignore.case = T) %>% 
               tibble(dartName=.) %>% 
               separate(dartName,c("germplasmName","dartID"),"_",extra = 'merge',remove = F)) %>% 
  group_by(germplasmName) %>% 
  slice(1) %>% 
  ungroup() %>% 
  rename(FullSampleName=dartName) %>% 
  mutate(OrigKeyFile="DArTseqLD", OriginOfSample="IITA") %>% 
  select(-dartID)
print(paste0(nrow(germNamesWithDArT)," germNames with DArT-only genos"))
# [1] "2401 germNames with DArT-only genos"


# first, filter to just program-DNAorigin matches
germNamesWithGenos<-dbdata %>% 
  select(programName,germplasmName) %>% 
  distinct %>% 
  left_join(gbs2phenoMaster) %>% 
  filter(!is.na(FullSampleName))
print(paste0(nrow(germNamesWithGenos)," germNames with GBS genos"))
# [1] "9323 germNames with GBS genos"

# program-germNames with locally sourced GBS samples
germNamesWithGenos_HasLocalSourcedGBS<-germNamesWithGenos %>% 
  filter(programName==OriginOfSample) %>% 
  select(programName,germplasmName) %>% 
  semi_join(germNamesWithGenos,.) %>% 
  group_by(programName,germplasmName) %>% # select one DNA per germplasmName per program
  slice(1) %>% ungroup() 
print(paste0(nrow(germNamesWithGenos_HasLocalSourcedGBS)," germNames with local GBS genos"))
# [1] "8257 germNames with local GBS genos"

# the rest (program-germNames) with GBS but coming from a different breeding program
germNamesWithGenos_NoLocalSourcedGBS<-germNamesWithGenos %>% 
  filter(programName==OriginOfSample) %>% 
  select(programName,germplasmName) %>% 
  anti_join(germNamesWithGenos,.) %>% 
  # select one DNA per germplasmName per program
  group_by(programName,germplasmName) %>% 
  slice(1) %>% ungroup() 
print(paste0(nrow(germNamesWithGenos_NoLocalSourcedGBS)," germNames without local GBS genos"))
# [1] "167 germNames without local GBS genos"

genosForPhenos<-bind_rows(germNamesWithGenos_HasLocalSourcedGBS,
                        germNamesWithGenos_NoLocalSourcedGBS) %>% 
  bind_rows(germNamesWithDArT)

print(paste0(nrow(genosForPhenos)," total germNames with genos either GBS or DArT"))
# [1] "10825 total germNames with genos either GBS or DArT"

dbdata %<>% 
    left_join(genosForPhenos) 

# Create a new identifier, GID
## Equals the value SNP data name (FullSampleName) 
## else germplasmName if no SNP data
dbdata %<>% 
  mutate(GID=ifelse(is.na(FullSampleName),germplasmName,FullSampleName))
```

```{r}
#  going to check against SNP data
# snps<-readRDS(file=url(paste0("ftp://ftp.cassavabase.org/marnin_datasets/NGC_BigData/",
#                               "DosageMatrix_RefPanelAndGSprogeny_ReadyForGP_73019.rds")))
# rownames_snps<-rownames(snps); rm(snps); gc()
# # current matches to SNP data
# dbdata %>% 
#   distinct(GID,germplasmName,FullSampleName) %>% 
#   semi_join(tibble(GID=rownames_snps)) %>% nrow() # 10707
# dbdata %>% 
#   distinct(GID,germplasmName,FullSampleName) %>% 
#   semi_join(tibble(GID=rownames_snps)) %>% 
#   filter(grepl("TMS13|2013_",GID,ignore.case = F)) %>% nrow() # 2424 TMS13
# dbdata %>% 
#   distinct(GID,germplasmName,FullSampleName) %>% 
#   semi_join(tibble(GID=rownames_snps)) %>% 
#   filter(grepl("TMS14",GID,ignore.case = F)) %>% nrow() # 2236 TMS14
# dbdata %>% 
#   distinct(GID,germplasmName,FullSampleName) %>% 
#   semi_join(tibble(GID=rownames_snps)) %>% 
#   filter(grepl("TMS15",GID,ignore.case = F)) %>% nrow() # 2287 TMS15
# dbdata %>% 
#   distinct(GID,germplasmName,FullSampleName) %>% 
#   semi_join(tibble(GID=rownames_snps)) %>% 
#   filter(grepl("TMS18",GID,ignore.case = F)) %>% nrow() # 2401 TMS18
```

# [User input] Choose locations

*WARNING: User input required!* If I had preselected locations before downloading, this wouldn't have been necessary. 

Based on previous locations used for IITA analysis, but adding based on list of locations used in IYR's trial list `data/2019_GS_PhenoUpload.csv`: "Ago-Owu" wasn't used last year.
```{r}
dbdata %<>% 
  filter(locationName %in% c("Abuja","Ago-Owu","Ibadan","Ikenne","Ilorin","Jos","Kano",
                             "Malam Madori","Mokwa","Ubiaja","Umudike","Warri","Zaria"))
nrow(dbdata) # [1] 432408
```


# Output "cleaned" file
```{r}
saveRDS(dbdata,file=here::here("output","IITA_CleanedTrialData_2020Dec03.rds"))
```

# Next step

2. [Get BLUPs combining all trial data](08-GetBLUPs_Dec2020.html): Detect experimental designs, Combine data from all trait-trials to get BLUPs for downstream genomic predictions. 
