---
title: "Genomic predictions"
author: "Marnin Wolfe"
date: "2020-Dec-03"
output: 
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = F, tidy = T)
```

# Previous step

2. [Get BLUPs combining all trial data](08-GetBLUPs_Dec2020.html): Detect experimental designs, Combine data from all trait-trials to get BLUPs for downstream genomic predictions. 

# Current Step

3. [Genomic prediction](09-GetGBLUPs_Dec2020.html): Predict **GETGV** specifically, for all selection candidates using all available data.

# Prediction using two-stages

## Set-up
```{bash, eval=F}
# activate multithread OpenBLAS 
export OMP_NUM_THREADS=44
```

```{r, eval=F}
library(tidyverse); library(magrittr); 
source(here::here("code","gsFunctions.R"))

A<-readRDS(file=here::here("output","Kinship_A_IITA_2020Sep16.rds"))
D<-readRDS(file=here::here("output","Kinship_D_IITA_2020Sep16.rds"))
AD<-readRDS(file=here::here("output","Kinship_AD_IITA_2020Sep16.rds"))

# BLUPs from the 2 stage procedure
# (stage 1 of 2) using the 2019 procedure
blups<-readRDS(file=here::here("output","iita_blupsForModelTraining_twostage_asreml_2020Dec03.rds")) %>% 
  dplyr::select(Trait,blups) %>% 
  unnest(blups) %>% 
  filter(GID %in% rownames(A)) %>% 
  nest(TrainingData=-Trait)
```

## Prediction

cbsurobbins (112 cores; 512GB) 

Model ADE
```{r, eval=F}
options(future.globals.maxSize= 3000*1024^2)
predModelADE<-runGenomicPredictions(blups,modelType="ADE",grms=list(A=A,D=D,AD=AD),gid="GID",ncores=14)
saveRDS(predModelADE,file = here::here("output","genomicPredictions_ModelADE_twostage_IITA_2020Dec03.rds"))
```

## Write GEBVs
```{r, eval=F}
library(tidyverse); library(magrittr);
predModelADE<-readRDS(file = here::here("output","genomicPredictions_ModelADE_twostage_IITA_2020Dec03.rds"))
traits<-c("MCMDS","DM","PLTHT","BRNHT1","BRLVLS","HI",
          "logDYLD", # <-- logDYLD now included. 
          "logFYLD","logTOPYLD","logRTNO","TCHART","LCHROMO","ACHROMO","BCHROMO")
```

```{r, eval=F}
## Format and write GETGV
predModelADE %>% 
  dplyr::select(Trait,genomicPredOut) %>% 
  unnest(genomicPredOut) %>% 
  select(-varcomps) %>% 
  unnest(gblups) %>% 
  select(Trait,GID,GETGV) %>% 
  spread(Trait,GETGV) %>% 
  mutate(GeneticGroup=NA,
         GeneticGroup=ifelse(grepl("TMS18",GID,ignore.case = T),"TMS18",
                             ifelse(grepl("TMS15",GID,ignore.case = T),"TMS15",
                                    ifelse(grepl("TMS14",GID,ignore.case = T),"TMS14",
                                           ifelse(grepl("TMS13|2013_",GID,ignore.case = T),"TMS13","GGetc"))))) %>% 
  select(GeneticGroup,GID,any_of(traits)) %>% arrange(desc(GeneticGroup)) %>% 
  write.csv(., file = here::here("output","GETGV_IITA_ModelADE_twostage_IITA_2020Dec03.csv"), row.names = F)
```

# Next step

6. [Results](06-Results.html#rate_of_gain)
